---
title: "Week 1: Cohort Assembly"
author: "Eric Xie"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

## Building the first AI-READI cohort table

### Setup

```{r}
library(tidyverse)

participants <- read_tsv("data/participants.tsv", show_col_types = FALSE)
cgm_manifest <- read_tsv("data/wearable_blood_glucose.tsv", 
                         show_col_types = FALSE)

cohort_table <- participants %>%
  transmute(
    person_id,
    age,
    study_group,
    clinical_site,
    recommended_split,
    cgm_availability = wearable_blood_glucose,
    wearable_availability = wearable_activity_monitor,
    clinical_data_availability = clinical_data
  ) %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

```

### Cohort Table (v3-mini)

- Combining the data sets of participant.tsv and manifest.tsv (in wearable_blood_glucose)
- ID variable "person_id" for joins

```{r}
knitr::kable(cohort_table)
```

## Tables and Figures

### Table 1: Participant Counts by Study Group
```{r}
table1 <- participants %>%
  count(study_group)

knitr::kable(table1)
```

This produces a summary table of how many participants are in each study group. Participants who are insulin dependent are noticeably less than other categories, within the scope of this data set.

### Table 2: Participant Counts by Study Group and Clinical Site
```{r}
table2 <- participants %>%
  count(study_group, clinical_site)

knitr::kable(table2)
```

This produces a summary table of how many participants are in each study group at each different clinical site. The number of participants are spread out relatively uniformly throughout the three clinic sites. 

### Table 3: Data Availability Summary
```{r}
# Table 3: data availability summary
table3 <- tibble(
  category = c(
    "CGM data",
    "Wearable data",
    "Clinical data",
    "CGM plus wearable data",
    "CGM plus wearable plus clinical data"
  ),
  count = c(
    sum(participants$wearable_blood_glucose == TRUE),
    sum(participants$wearable_activity_monitor == TRUE),
    sum(participants$clinical_data == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE 
        & participants$clinical_data == TRUE)
  )
)

knitr::kable(table3)
```

The count for each of the categories is the same (100), showing no variation. Therefore, all the participants in the study have complete data across all categories in this subset.

### Modality Availability Heatmap

```{r}
library(tidyverse)

modality_long <- participants %>%
  select(person_id, cardiac_ecg, clinical_data, environment,
         retinal_flio, retinal_oct, retinal_octa, retinal_photography,
         wearable_activity_monitor, wearable_blood_glucose) %>%
  pivot_longer(-person_id, names_to = "modality", values_to = "available")

ggplot(modality_long, aes(x = modality, y = factor(person_id), fill = available)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "#f0f0f0")) +
  labs(title = "Figure 1: Modality Availability Heatmap",
       x = "Modality", y = "Participant", fill = "Available") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),   # 100 IDs won't fit legibly on y-axis
        axis.text.x = element_text(angle = 45, hjust = 1))
```

When visualized, the participants' data is complete across all modalities in this subset. There is no missing modalities to present or analyze.

### Average CGM Glucose by Study Group (Boxplot)

```{r}
library(tidyverse)

cohort_glucose <- participants %>%
  select(person_id, study_group) %>%
  left_join(
    cgm_manifest %>% select(person_id, average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

ggplot(cohort_glucose, aes(x = study_group, y = average_cgm_glucose, fill = study_group)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  labs(title = "Average CGM Glucose by Study Group",
       x = "Study Group", y = "Average CGM Glucose (mg/dL)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

## Final Questions





