---
title: "Week 1: Cohort Assembly"
author: "Eric Xie"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

## Building the first AI-READI cohort table

### Setup

```{r}
library(tidyverse)

participants <- read_tsv("data/participants_mini.tsv", show_col_types = FALSE)
participants_full <- read_tsv("data/participants_full.tsv", 
                              show_col_types = FALSE)

cgm_manifest_full <- read_tsv("data/wearable_blood_glucose_full.tsv", 
                         show_col_types = FALSE)
cgm_manifest <- read_tsv("data/wearable_blood_glucose_mini.tsv", 
                         show_col_types = FALSE)

cohort_table <- participants %>%
  transmute(
    person_id,
    age,
    study_group,
    clinical_site,
    recommended_split,
    cgm_availability = wearable_blood_glucose,
    wearable_availability = wearable_activity_monitor,
    clinical_data_availability = clinical_data
  ) %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

```

### Cohort Table (v3-mini)

- Combining the data sets of participant.tsv and manifest.tsv (in wearable_blood_glucose)
- ID variable "person_id" for joins

```{r}
knitr::kable(cohort_table)
```

## Tables and Figures (mini version)

### Table 1: Participant Counts by Study Group
```{r}
table1 <- participants %>%
  count(study_group)

knitr::kable(table1)
```

This produces a summary table of how many participants are in each study group. Participants who are insulin dependent are noticeably less than other categories, within the scope of this data set.

### Table 2: Participant Counts by Study Group and Clinical Site
```{r}
table2 <- participants %>%
  count(study_group, clinical_site)

knitr::kable(table2)
```

This produces a summary table of how many participants are in each study group at each different clinical site. The number of participants are spread out relatively uniformly throughout the three clinic sites. 

### Table 3: Data Availability Summary
```{r}
table3 <- tibble(
  category = c(
    "CGM data",
    "Wearable data",
    "Clinical data",
    "CGM plus wearable data",
    "CGM plus wearable plus clinical data"
  ),
  count = c(
    sum(participants$wearable_blood_glucose == TRUE),
    sum(participants$wearable_activity_monitor == TRUE),
    sum(participants$clinical_data == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE 
        & participants$clinical_data == TRUE)
  )
)

knitr::kable(table3)
```

The count for each of the categories is the same (100), showing no variation. Therefore, all the participants in the study have complete data across all categories in this subset.

### Modality Availability Heatmap

```{r}
library(tidyverse)

modality_long <- participants %>%
  select(person_id, cardiac_ecg, clinical_data, environment,
         retinal_flio, retinal_oct, retinal_octa, retinal_photography,
         wearable_activity_monitor, wearable_blood_glucose) %>%
  pivot_longer(-person_id, names_to = "modality", values_to = "available")

ggplot(modality_long, aes(x = modality, y = factor(person_id), fill = available)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "#f0f0f0")) +
  labs(title = "Figure 1: Modality Availability Heatmap",
       x = "Modality", y = "Participant", fill = "Available") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),   # 100 IDs won't fit legibly on y-axis
        axis.text.x = element_text(angle = 45, hjust = 1))
```

When visualized, the participants' data is complete across all modalities in this subset. There is no missing modalities to present or analyze.

### Average CGM Glucose by Study Group (Boxplot)

```{r}
library(tidyverse)

cohort_glucose <- participants %>%
  select(person_id, study_group) %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

ggplot(cohort_glucose, 
       aes(x = study_group, y = average_cgm_glucose, fill = study_group)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  labs(title = "Average CGM Glucose by Study Group",
       x = "Study Group", y = "Average CGM Glucose (mg/dL)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

This boxplot shows the average CGM Glucose levels by study group (and their 25th and 75th percentile ranges). 

- The insulin-dependent study group noticeably has much higher average CGM glucose levels 
- The group "oral_medication_and_or_non_insulin_injectable_medication_controlled" has the most variation / range of values
- The insulin-dependent group's values are very clustered towards the median, except two outliers
- The variations for "healthy" and "pre_diabetes_lifestyle_controlled" are roughly the same

Note: In geom_boxplot(),the whisker extends to the most extreme data point that still falls within 1.5 × IQR of the box edge, where IQR is the height of the box itself (Q3 − Q1). Anything past that boundary is classified as an outlier.

## Tables and Figures (full v3 version)

### Table 1: Participant Counts by Study Group
```{r}
table1_full <- participants_full %>%
  count(study_group)

knitr::kable(table1_full)
```

### Table 2: Participant Counts by Study Group and Clinical Site
```{r}
table2_full <- participants_full %>%
  count(study_group, clinical_site)

knitr::kable(table2_full)
```

### Table 3: Data Availability Summary
```{r}
table3_full <- tibble(
  category = c(
    "CGM data",
    "Wearable data",
    "Clinical data",
    "CGM plus wearable data",
    "CGM plus wearable plus clinical data"
  ),
  count = c(
    sum(participants_full$wearable_blood_glucose == TRUE),
    sum(participants_full$wearable_activity_monitor == TRUE),
    sum(participants_full$clinical_data == TRUE),
    sum(participants_full$wearable_blood_glucose == TRUE 
        & participants_full$wearable_activity_monitor == TRUE),
    sum(participants_full$wearable_blood_glucose == TRUE 
        & participants_full$wearable_activity_monitor == TRUE 
        & participants_full$clinical_data == TRUE)
  )
)

knitr::kable(table3_full)
```

### Average CGM Glucose by Study Group (Boxplot)

```{r}
library(tidyverse)

cohort_glucose_full <- participants_full %>%
  select(person_id, study_group) %>%
  left_join(
    cgm_manifest_full %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

ggplot(cohort_glucose_full, 
       aes(x = study_group, y = average_cgm_glucose, fill = study_group)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  labs(title = "Average CGM Glucose by Study Group",
       x = "Study Group", y = "Average CGM Glucose (mg/dL)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

## Mini Versus Full Dataset Comparison

### How many participants are in the mini dataset versus the full dataset?

```{r}
library(tidyverse)

dataset_sizes <- tibble(
  dataset = c("Mini", "Full"),
  n_participants = c(100, 2280)
)

ggplot(dataset_sizes, aes(x = dataset, y = n_participants, fill = dataset)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = n_participants), vjust = -0.5, size = 5) +
  scale_fill_manual(values = c("Mini" = "#85B7EB", "Full" = "#185FA5")) +
  labs(title = "Participants: Mini vs Full Dataset",
       x = NULL, y = "Number of Participants") +
  theme_minimal() +
  theme(legend.position = "none")
```

The mini dataset has 100 participants, while the full set has 2,280. 

### Are the study groups represented similarly?

The study groups are mostly represented similarly, with the exception of the insulin-dependent group, which has significantly less participants.

### Are the clinical sites represented similarly?

Yes, the clinical sites have roughly equal representation throughout the study.

### Is modality availability similar or different?

The mini dataset has every modality available, however, for the full data set, some are missing (though it's close to 100%).

### Did the same code run on both datasets after changing only the data path?

Yes, they mostly did, with the exception of the heatmap code, whose graphics were blurred by the overly tiny marks due to the large sample population.

## Join and Data Quality Checks (mini dataset)

### Number of rows before and after each join

```{r}
rows_before_participants <- nrow(participants)
rows_before_cgm <- nrow(cgm_manifest)

cohort_glucose <- participants %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

rows_after_join <- nrow(cohort_glucose)

qc_rows <- tibble(
  step = c("participants.tsv (before join)",
           "wearable_blood_glucose.tsv (before join)",
           "after join"),
  row_count = c(rows_before_participants, rows_before_cgm, rows_after_join)
)

knitr::kable(qc_rows)
```

### Number of unique participants before and after each join

```{r}
unique_before_participants <- n_distinct(participants$person_id)
unique_before_cgm <- n_distinct(cgm_manifest$person_id)

cohort_glucose <- participants %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

unique_after_join <- n_distinct(cohort_glucose$person_id)

qc_unique <- tibble(
  step = c("participants.tsv (before join)",
           "wearable_blood_glucose.tsv (before join)",
           "after join"),
  unique_participants = c(unique_before_participants, 
                          unique_before_cgm, unique_after_join)
)

knitr::kable(qc_unique)
```

### Whether any participant IDs are duplicated

```{r}
dup_participants <- participants %>%
  count(person_id) %>%
  filter(n > 1)

dup_cgm <- cgm_manifest %>%
  count(person_id) %>%
  filter(n > 1)

dup_after_join <- cohort_glucose %>%
  count(person_id) %>%
  filter(n > 1)

qc_duplicates <- tibble(
  step = c("participants.tsv", "wearable_blood_glucose.tsv", "after join"),
  n_duplicated_ids = c(nrow(dup_participants), 
                       nrow(dup_cgm), nrow(dup_after_join))
)

knitr::kable(qc_duplicates)
```

### Whether any participants are lost unexpectedly
```{r}
# IDs present in participants.tsv but missing from the CGM manifest
missing_from_cgm <- setdiff(participants$person_id, cgm_manifest$person_id)

# IDs present in the CGM manifest but missing from participants.tsv
missing_from_participants <- setdiff(cgm_manifest$person_id, 
                                     participants$person_id)

# after the join, check for any row where the join produced NA
# (i.e., a participant with no matching CGM record)
lost_after_join <- cohort_glucose %>%
  filter(is.na(average_cgm_glucose)) %>%
  pull(person_id)

qc_lost <- tibble(
  check = c("In participants.tsv but not in CGM manifest",
            "In CGM manifest but not in participants.tsv",
            "Participants with no CGM match after join"),
  n_ids = c(length(missing_from_cgm),
            length(missing_from_participants),
            length(lost_after_join))
)

knitr::kable(qc_lost)
```

### Which variables have the most missing values

```{r}
missing_summary <- cohort_glucose %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  arrange(desc(n_missing))

knitr::kable(missing_summary)
```

## Analysis Cohort Decisions

### Code

```{r}
cgm_only_cohort <- participants %>%
  filter(wearable_blood_glucose == TRUE)

cgm_plus_wearable_cohort <- participants %>%
  filter(wearable_blood_glucose == TRUE, wearable_activity_monitor == TRUE)

n_cgm_only <- nrow(cgm_only_cohort)
n_cgm_plus_wearable <- nrow(cgm_plus_wearable_cohort)

tibble(
  cohort = c("CGM-only", "CGM and wearable"),
  n_participants = c(n_cgm_only, n_cgm_plus_wearable)
) %>%
  knitr::kable()
```

### CGM-only cohort

In the cohort, 100 participants are included, representing the study groups of oral medication/non-insulin controlled, healthy, pre-diabetes lifestyle-controlled, and insulin-dependent. Since each of the 100 participants have full availability of the modalities, any scientific modality-related question could be asked. However, the cohort is limited in that the mini dataset was selected from the full data set purposefully without any missingness of data, an element that cannot be explained by this cohort. 

### CGM plus wearable cohort

In the CGM plus wearable cohort, 100 participants are included, representing the same study groups as the CGM-only cohort. It's potential scientific inquiries and limitations are also similar, but this cohort might be able to provide more complete insight into daily glucose monitoring.

## Interpretation

### Missing Data

```{r}
missing_summary_full <- participants_full %>%
  summarise(across(c(cardiac_ecg, clinical_data, environment,
                      retinal_flio, retinal_oct, retinal_octa,
                      retinal_photography, wearable_activity_monitor,
                      wearable_blood_glucose),
                    ~ sum(. == FALSE))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(
    total = nrow(participants_full),
    pct_missing = round(100 * n_missing / total, 1)
  ) %>%
  select(variable, total, n_missing, pct_missing) %>%
  arrange(desc(n_missing))

knitr::kable(missing_summary_full)
```

### What does the missing-data pattern tell us about which scientific questions are possible in this dataset?

Overall, the missing data in the full data set limits which modalities can be analyzed, whether it be alone or jointly with other modalities. Noticeably, the variable "retinal_flio" has the most missing data out of all other variables, a specialized retinal imaging technique that is perhaps implemented less than the standard retinal OCT, which has more complete data. Therefore, although the data set could support many other inquiries such as CGM or clinical data, but questions that require FLIO would be limited in what conclusions they could result in. Furthermore, the missing data not only points to the lack of completeness, but also underlying sample biases that could determine how the data set is formed. 

### Why is it useful to develop code on the mini dataset but check scientific conclusions on the full dataset?

Developing code on the mini dataset could allow for increased speed and smaller-scale iterations which could aid debugging during the initial code-writing process. The smaller data analysis could point towards potential trends in the entire full data set, but to verify such trends and conclusions the full dataset must be evaluated for completeness and to mitigate sampling bias. 

## Final Questions

### The mini dataset doesn't have any missing data for each participant, but the full dataset does for some. Was the mini data set deliberately constructed this way such that it would be complete, or was it created using another standard or randomness?




