---
title: "Week 1: Cohort Assembly"
author: "Eric Xie"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

## Building the first AI-READI cohort table

### Setup

```{r}
library(tidyverse)

participants <- read_tsv("data/participants.tsv", show_col_types = FALSE)
cgm_manifest <- read_tsv("data/wearable_blood_glucose.tsv", 
                         show_col_types = FALSE)

cohort_table <- participants %>%
  transmute(
    person_id,
    age,
    study_group,
    clinical_site,
    recommended_split,
    cgm_availability = wearable_blood_glucose,
    wearable_availability = wearable_activity_monitor,
    clinical_data_availability = clinical_data
  ) %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

```

### Cohort Table (v3-mini)

- Combining the data sets of participant.tsv and manifest.tsv (in wearable_blood_glucose)
- ID variable "person_id" for joins

```{r}
knitr::kable(cohort_table)
```

## Tables and Figures

### Table 1: Participant Counts by Study Group
```{r}
table1 <- participants %>%
  count(study_group)

knitr::kable(table1)
```

### Table 2: Participant Counts by Study Group and Clinical Site
```{r}
table2 <- participants %>%
  count(study_group, clinical_site)

knitr::kable(table2)
```

### Table 3: Data Availability Summary
```{r}
# Table 3: data availability summary
table3 <- tibble(
  category = c(
    "CGM data",
    "Wearable data",
    "Clinical data",
    "CGM plus wearable data",
    "CGM plus wearable plus clinical data"
  ),
  count = c(
    sum(participants$wearable_blood_glucose == TRUE),
    sum(participants$wearable_activity_monitor == TRUE),
    sum(participants$clinical_data == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE 
        & participants$clinical_data == TRUE)
  )
)

knitr::kable(table3)
```

## Final Questions

1. The meeting notes mentioned an ID variable you use for joins, but I'm unclear on what this is.




