---
title: "Week 1: Cohort Assembly"
author: "Eric Xie"
format:
  html:
    toc: true
    toc-location: left
    number-sections: true
---

## Building the first AI-READI cohort table

### Setup

```{r}
library(tidyverse)

participants <- read_tsv("data/participants_mini.tsv", show_col_types = FALSE)
participants_full <- read_tsv("data/participants_full.tsv", 
                              show_col_types = FALSE)

cgm_manifest_full <- read_tsv("data/wearable_blood_glucose_full.tsv", 
                         show_col_types = FALSE)
cgm_manifest <- read_tsv("data/wearable_blood_glucose_mini.tsv", 
                         show_col_types = FALSE)

cohort_table <- participants %>%
  transmute(
    person_id,
    age,
    study_group,
    clinical_site,
    recommended_split,
    cgm_availability = wearable_blood_glucose,
    wearable_availability = wearable_activity_monitor,
    clinical_data_availability = clinical_data
  ) %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

```

### Cohort Table (v3-mini)

- Combining the data sets of participant.tsv and manifest.tsv (in wearable_blood_glucose)
- ID variable "person_id" for joins

```{r}
knitr::kable(cohort_table)
```

## Tables and Figures (mini version)

### Table 1: Participant Counts by Study Group
```{r}
table1 <- participants %>%
  count(study_group)

knitr::kable(table1)
```

This produces a summary table of how many participants are in each study group. Participants who are insulin dependent are noticeably less than other categories, within the scope of this data set.

### Table 2: Participant Counts by Study Group and Clinical Site
```{r}
table2 <- participants %>%
  count(study_group, clinical_site)

knitr::kable(table2)
```

This produces a summary table of how many participants are in each study group at each different clinical site. The number of participants are spread out relatively uniformly throughout the three clinic sites. 

### Table 3: Data Availability Summary
```{r}
# Table 3: data availability summary
table3 <- tibble(
  category = c(
    "CGM data",
    "Wearable data",
    "Clinical data",
    "CGM plus wearable data",
    "CGM plus wearable plus clinical data"
  ),
  count = c(
    sum(participants$wearable_blood_glucose == TRUE),
    sum(participants$wearable_activity_monitor == TRUE),
    sum(participants$clinical_data == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE),
    sum(participants$wearable_blood_glucose == TRUE 
        & participants$wearable_activity_monitor == TRUE 
        & participants$clinical_data == TRUE)
  )
)

knitr::kable(table3)
```

The count for each of the categories is the same (100), showing no variation. Therefore, all the participants in the study have complete data across all categories in this subset.

### Modality Availability Heatmap

```{r}
library(tidyverse)

modality_long <- participants %>%
  select(person_id, cardiac_ecg, clinical_data, environment,
         retinal_flio, retinal_oct, retinal_octa, retinal_photography,
         wearable_activity_monitor, wearable_blood_glucose) %>%
  pivot_longer(-person_id, names_to = "modality", values_to = "available")

ggplot(modality_long, aes(x = modality, y = factor(person_id), fill = available)) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_manual(values = c("TRUE" = "steelblue", "FALSE" = "#f0f0f0")) +
  labs(title = "Figure 1: Modality Availability Heatmap",
       x = "Modality", y = "Participant", fill = "Available") +
  theme_minimal() +
  theme(axis.text.y = element_blank(),   # 100 IDs won't fit legibly on y-axis
        axis.text.x = element_text(angle = 45, hjust = 1))
```

When visualized, the participants' data is complete across all modalities in this subset. There is no missing modalities to present or analyze.

### Average CGM Glucose by Study Group (Boxplot)

```{r}
library(tidyverse)

cohort_glucose <- participants %>%
  select(person_id, study_group) %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

ggplot(cohort_glucose, 
       aes(x = study_group, y = average_cgm_glucose, fill = study_group)) +
  geom_boxplot() +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  labs(title = "Average CGM Glucose by Study Group",
       x = "Study Group", y = "Average CGM Glucose (mg/dL)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1),
        legend.position = "none")
```

This boxplot shows the average CGM Glucose levels by study group (and their 25th and 75th percentile ranges). 
- The insulin-dependent study group noticeably has much higher average CGM glucose levels 
- The group "oral_medication_and_or_non_insulin_injectable_medication_controlled" has the most variation / range of values
- The insulin-dependent group's values are very clustered towards the median, except two outliers
- The variations for "healthy" and "pre_diabetes_lifestyle_controlled" are roughly the same

## Tables and Figures (mini version)

### Table 1: Participant Counts by Study Group
```{r}

```

## Mini Versus Full Dataset Comparison


## Join and Data Quality Checks

### Number of rows before and after each join

```{r}
rows_before_participants <- nrow(participants)
rows_before_cgm <- nrow(cgm_manifest)

cohort_glucose <- participants %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

rows_after_join <- nrow(cohort_glucose)

qc_rows <- tibble(
  step = c("participants.tsv (before join)",
           "wearable_blood_glucose.tsv (before join)",
           "after join"),
  row_count = c(rows_before_participants, rows_before_cgm, rows_after_join)
)

knitr::kable(qc_rows)
```

### Number of unique participants before and after each join

```{r}
unique_before_participants <- n_distinct(participants$person_id)
unique_before_cgm <- n_distinct(cgm_manifest$person_id)

cohort_glucose <- participants %>%
  left_join(
    cgm_manifest %>% select(person_id, 
                            average_cgm_glucose = average_glucose_level_mg_dl),
    by = "person_id"
  )

unique_after_join <- n_distinct(cohort_glucose$person_id)

qc_unique <- tibble(
  step = c("participants.tsv (before join)",
           "wearable_blood_glucose.tsv (before join)",
           "after join"),
  unique_participants = c(unique_before_participants, 
                          unique_before_cgm, unique_after_join)
)

knitr::kable(qc_unique)
```

### Whether any participant IDs are duplicated

```{r}
dup_participants <- participants %>%
  count(person_id) %>%
  filter(n > 1)

dup_cgm <- cgm_manifest %>%
  count(person_id) %>%
  filter(n > 1)

dup_after_join <- cohort_glucose %>%
  count(person_id) %>%
  filter(n > 1)

qc_duplicates <- tibble(
  step = c("participants.tsv", "wearable_blood_glucose.tsv", "after join"),
  n_duplicated_ids = c(nrow(dup_participants), 
                       nrow(dup_cgm), nrow(dup_after_join))
)

knitr::kable(qc_duplicates)
```

### Whether any participants are lost unexpectedly
```{r}
# IDs present in participants.tsv but missing from the CGM manifest
missing_from_cgm <- setdiff(participants$person_id, cgm_manifest$person_id)

# IDs present in the CGM manifest but missing from participants.tsv
missing_from_participants <- setdiff(cgm_manifest$person_id, 
                                     participants$person_id)

# after the join, check for any row where the join produced NA
# (i.e., a participant with no matching CGM record)
lost_after_join <- cohort_glucose %>%
  filter(is.na(average_cgm_glucose)) %>%
  pull(person_id)

qc_lost <- tibble(
  check = c("In participants.tsv but not in CGM manifest",
            "In CGM manifest but not in participants.tsv",
            "Participants with no CGM match after join"),
  n_ids = c(length(missing_from_cgm),
            length(missing_from_participants),
            length(lost_after_join))
)

knitr::kable(qc_lost)
```

### Which variables have the most missing values

```{r}
missing_summary <- cohort_glucose %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  arrange(desc(n_missing))

knitr::kable(missing_summary)
```

## Analysis Cohort Decisions
```{r}
cgm_only_cohort <- participants %>%
  filter(wearable_blood_glucose == TRUE)

cgm_plus_wearable_cohort <- participants %>%
  filter(wearable_blood_glucose == TRUE, wearable_activity_monitor == TRUE)

n_cgm_only <- nrow(cgm_only_cohort)
n_cgm_plus_wearable <- nrow(cgm_plus_wearable_cohort)

tibble(
  cohort = c("CGM-only", "CGM and wearable"),
  n_participants = c(n_cgm_only, n_cgm_plus_wearable)
) %>%
  knitr::kable()
```

### CGM-only cohort: participants with usable CGM data



## Final Questions





