---
title: "Week 2: CGM Processing and Individual Glucose Patterns"
author: "Eric Xie"
format:
  html:
    toc: true
    toc-location: left
---

## CGM Cohort
```{r}
library(tidyverse)

participants_mini <- read_tsv("data/participants_mini.tsv")
cgm_cohort <- participants_mini %>%
  filter(wearable_blood_glucose == TRUE) %>%
  pull(person_id)

manifest <- read_tsv("data/wearable_blood_glucose_mini.tsv")
manifest %>% pull(glucose_filepath) %>% head(3)
```

## Cleaning the CGM Data

##### Keeping Variables
```{r}
library(lubridate)
library(jsonlite)

json_path_1029 <- "data/dexcom_g6/1029/1029_DEX.json"
raw_1029 <- fromJSON(json_path_1029, simplifyVector = FALSE)

person_id_val   <- raw_1029$header$patient_id
timezone_abbrev <- raw_1029$header$timezone

cgm_1029 <- map_dfr(raw_1029$body$cgm, function(r) {
  tibble(
    person_id             = person_id_val,
    event_type            = r$event_type,
    time_utc              = r$effective_time_frame$time_interval$start_date_time,
    timezone_abbreviation = timezone_abbrev,
    glucose_value_original = r$blood_glucose$value,
    glucose_value_numeric  = as.numeric(r$blood_glucose$value),
    glucose_unit            = r$blood_glucose$unit
  )
})

glimpse(cgm_1029)
nrow(cgm_1029)
```

##### 1. Keep only estimated glucose value events (event_type == "EGV")
```{r}
cgm_1029 <- cgm_1029 %>%
  filter(event_type == "EGV")

nrow(cgm_1029)
```

##### 2. Parse timestamps ending in Z as UTC and save them as time_utc
```{r}
cgm_1029 <- cgm_1029 %>%
  mutate(time_utc = ymd_hms(time_utc, tz = "UTC"))

# cgm_1029$person_id is "AIREADI-1029"; participants_mini$person_id is bare 1029
cgm_1029 <- cgm_1029 %>%
  mutate(person_id_numeric = as.numeric(str_remove(person_id, "AIREADI-")))
```

##### 3. Infer the participant/device regional timezone from the JSON timezone label and check that it is consistent with the clinical site.
```{r}
tz_map <- tribble(
  ~timezone_abbreviation, ~iana_zone,           ~expected_region,
  "pst",                  "America/Los_Angeles", "Pacific",
  "pdt",                  "America/Los_Angeles", "Pacific",
  "mst",                  "America/Denver",       "Mountain",
  "mdt",                  "America/Denver",       "Mountain",
  "cst",                  "America/Chicago",      "Central",
  "cdt",                  "America/Chicago",      "Central",
  "est",                  "America/New_York",     "Eastern",
  "edt",                  "America/New_York",     "Eastern"
)

site_region_map <- tribble(
  ~clinical_site, ~site_region,
  "UW",           "Pacific",
  "UCSD",         "Pacific",
  "UAB",          "Central"
)

cgm_1029 <- cgm_1029 %>%
  left_join(tz_map, by = "timezone_abbreviation") %>%
  left_join(
    participants_mini %>% select(person_id, clinical_site),
    by = c("person_id_numeric" = "person_id")
  ) %>%
  left_join(site_region_map, by = "clinical_site") %>%
  mutate(
    tz_site_consistent = expected_region == site_region,
    tz_flag_for_review = is.na(iana_zone) | is.na(site_region) | !tz_site_consistent
  )

cgm_1029 %>%
  distinct(person_id, timezone_abbreviation, iana_zone, clinical_site, site_region, tz_site_consistent, tz_flag_for_review)
```

Nothing flagged for participant 1029. 

##### 4. Convert time_utc to time_local with the selected IANA timezone. Create date_local, hour_local, timezone_abbreviation, and is_dst. Let R apply daylight-saving-time rules automatically.

```{r}
participant_tz <- unique(cgm_1029$iana_zone)
stopifnot(length(participant_tz) == 1)  # sanity check — should be exactly one zone

cgm_1029 <- cgm_1029 %>%
  mutate(
    time_local = with_tz(time_utc, tzone = participant_tz),
    date_local = as_date(time_local),
    hour_local = hour(time_local),
    is_dst     = dst(time_local)
  )

cgm_1029 %>%
  select(time_utc, time_local, date_local, hour_local, is_dst) %>%
  head(10) %>%
  knitr::kable()
```

##### 6. Check for missing timestamps and glucose values
```{r}
missing_summary <- cgm_1029 %>%
  summarize(
    n_total                   = n(),
    n_missing_time_utc         = sum(is.na(time_utc)),
    n_missing_glucose_original = sum(is.na(glucose_value_original)),
    n_missing_glucose_numeric  = sum(is.na(glucose_value_numeric))
  )

missing_summary %>% knitr::kable()
```

##### 7. Define a duplicate as the same participant ID and UTC timestamp.

- If duplicate rows have the same glucose value, keep one copy.
- If duplicate rows have different glucose values, flag them for review and exclude them from the summary calculations.

```{r}
dup_groups <- cgm_1029 %>%
  count(person_id, time_utc, name = "n_readings") %>%
  filter(n_readings > 1)

nrow(dup_groups)
```



