---
title: "Week 2: CGM Processing and Individual Glucose Patterns"
author: "Eric Xie"
format:
  html:
    toc: true
    toc-location: left
---

## CGM Cohort
```{r}
library(tidyverse)

participants_mini <- read_tsv("data/participants_mini.tsv")
cgm_cohort <- participants_mini %>%
  filter(wearable_blood_glucose == TRUE) %>%
  pull(person_id)

manifest <- read_tsv("data/wearable_blood_glucose_mini.tsv")
manifest %>% pull(glucose_filepath) %>% head(3)
```

## Cleaning the CGM Data

##### Keeping Variables
```{r}
library(lubridate)
library(jsonlite)

json_path_1029 <- "data/dexcom_g6/1029/1029_DEX.json"
raw_1029 <- fromJSON(json_path_1029, simplifyVector = FALSE)

person_id_val   <- raw_1029$header$patient_id
timezone_abbrev <- raw_1029$header$timezone

cgm_1029 <- map_dfr(raw_1029$body$cgm, function(r) {
  tibble(
    person_id             = person_id_val,
    event_type            = r$event_type,
    time_utc              = r$effective_time_frame$time_interval$start_date_time,
    timezone_abbreviation = timezone_abbrev,
    glucose_value_original = r$blood_glucose$value,
    glucose_value_numeric  = as.numeric(r$blood_glucose$value),
    glucose_unit            = r$blood_glucose$unit
  )
})

glimpse(cgm_1029)
nrow(cgm_1029)
```

##### 1. Keep only estimated glucose value events (event_type == "EGV")
```{r}
cgm_1029 <- cgm_1029 %>%
  filter(event_type == "EGV")

nrow(cgm_1029)
```

##### 2. Parse timestamps ending in Z as UTC and save them as time_utc
```{r}
cgm_1029 <- cgm_1029 %>%
  mutate(time_utc = ymd_hms(time_utc, tz = "UTC"))

# cgm_1029$person_id is "AIREADI-1029"; participants_mini$person_id is bare 1029
cgm_1029 <- cgm_1029 %>%
  mutate(person_id_numeric = as.numeric(str_remove(person_id, "AIREADI-")))
```

##### 3. Infer the participant/device regional timezone from the JSON timezone label and check that it is consistent with the clinical site.
```{r}
tz_map <- tribble(
  ~timezone_abbreviation, ~iana_zone,           ~expected_region,
  "pst",                  "America/Los_Angeles", "Pacific",
  "pdt",                  "America/Los_Angeles", "Pacific",
  "mst",                  "America/Denver",       "Mountain",
  "mdt",                  "America/Denver",       "Mountain",
  "cst",                  "America/Chicago",      "Central",
  "cdt",                  "America/Chicago",      "Central",
  "est",                  "America/New_York",     "Eastern",
  "edt",                  "America/New_York",     "Eastern"
)

site_region_map <- tribble(
  ~clinical_site, ~site_region,
  "UW",           "Pacific",
  "UCSD",         "Pacific",
  "UAB",          "Central"
)

cgm_1029 <- cgm_1029 %>%
  left_join(tz_map, by = "timezone_abbreviation") %>%
  left_join(
    participants_mini %>% select(person_id, clinical_site),
    by = c("person_id_numeric" = "person_id")
  ) %>%
  left_join(site_region_map, by = "clinical_site") %>%
  mutate(
    tz_site_consistent = expected_region == site_region,
    tz_flag_for_review = is.na(iana_zone) | is.na(site_region) | !tz_site_consistent
  )

cgm_1029 %>%
  distinct(person_id, timezone_abbreviation, iana_zone, clinical_site, site_region, tz_site_consistent, tz_flag_for_review)
```

Nothing flagged for participant 1029. 

##### 4. Convert time_utc to time_local with the selected IANA timezone. Create date_local, hour_local, timezone_abbreviation, and is_dst. Let R apply daylight-saving-time rules automatically.

```{r}
participant_tz <- unique(cgm_1029$iana_zone)
stopifnot(length(participant_tz) == 1)  # sanity check — should be exactly one zone

cgm_1029 <- cgm_1029 %>%
  mutate(
    time_local = with_tz(time_utc, tzone = participant_tz),
    date_local = as_date(time_local),
    hour_local = hour(time_local),
    is_dst     = dst(time_local)
  )

cgm_1029 %>%
  select(time_utc, time_local, date_local, hour_local, is_dst) %>%
  head(10) %>%
  knitr::kable()
```

##### 6. Check for missing timestamps and glucose values
```{r}
missing_summary <- cgm_1029 %>%
  summarize(
    n_total                   = n(),
    n_missing_time_utc         = sum(is.na(time_utc)),
    n_missing_glucose_original = sum(is.na(glucose_value_original)),
    n_missing_glucose_numeric  = sum(is.na(glucose_value_numeric))
  )

missing_summary %>% knitr::kable()
```

##### 7. Define a duplicate as the same participant ID and UTC timestamp.

- If duplicate rows have the same glucose value, keep one copy.
- If duplicate rows have different glucose values, flag them for review and exclude them from the summary calculations.

```{r}
dup_groups <- cgm_1029 %>%
  count(person_id, time_utc, name = "n_readings") %>%
  filter(n_readings > 1)

nrow(dup_groups)

cgm_1029_clean <- cgm_1029 %>%
  distinct(person_id, time_utc, .keep_all = TRUE)

nrow(cgm_1029)
nrow(cgm_1029_clean)
```

##### 8. Confirm glucose unit is mg/dL
```{r}
unit_check <- cgm_1029_clean %>%
  count(glucose_unit, name = "n_readings")

unit_check %>% knitr::kable()
```

##### 9. Flag unusual glucose values (< 40 or > 400 mg/dL)
```{r}
cgm_1029_clean <- cgm_1029_clean %>%
  mutate(unusual_value_flag = glucose_value_numeric < 40 | glucose_value_numeric > 400)

n_unusual <- sum(cgm_1029_clean$unusual_value_flag, na.rm = TRUE)
n_unusual

cgm_1029_valid_readings <- cgm_1029_clean %>%
  filter(!unusual_value_flag)

nrow(cgm_1029_clean)
nrow(cgm_1029_valid_readings)
```

##### 10. Create the clean reading-level CGM table
```{r}
cgm_1029_reading_level <- cgm_1029_valid_readings %>%
  mutate(person_id = person_id_numeric) %>%
  select(
    person_id,
    event_type,
    time_utc,
    timezone_abbreviation,
    time_local,
    date_local,
    hour_local,
    is_dst,
    glucose_value_original,
    glucose_value_numeric,
    glucose_unit
  )

nrow(cgm_1029)
nrow(cgm_1029_reading_level)

cgm_1029_reading_level %>%
  head(10) %>%
  knitr::kable()
```

##### 11. Count readings per local day and flag valid days
```{r}
cgm_1029_day_counts <- cgm_1029_reading_level %>%
  count(person_id, date_local, name = "n_readings") %>%
  mutate(valid_day = n_readings >= 202)

cgm_1029_day_counts %>% knitr::kable()
```

##### 12. Flag any date that crosses a DST transition
```{r}
dst_crossing_days <- cgm_1029_reading_level %>%
  group_by(person_id, date_local) %>%
  summarize(
    n_distinct_dst = n_distinct(is_dst),
    .groups = "drop"
  ) %>%
  filter(n_distinct_dst > 1)

nrow(dst_crossing_days)
dst_crossing_days %>% knitr::kable()
```

## Summary Tables

##### Participant-Level Summary
```{r}
# Restrict to cleaned readings that fall on valid local days only
cgm_1029_valid_days_readings <- cgm_1029_reading_level %>%
  left_join(
    cgm_1029_day_counts %>% select(person_id, date_local, valid_day),
    by = c("person_id", "date_local")
  ) %>%
  filter(valid_day)

nrow(cgm_1029_reading_level)         # all cleaned readings
nrow(cgm_1029_valid_days_readings)   # cleaned readings from valid days only
```

```{r}
cgm_1029_participant_summary <- cgm_1029_valid_days_readings %>%
  summarize(
    person_id      = first(person_id),
    n_readings     = n(),
    n_valid_days   = n_distinct(date_local),
    mean_glucose   = mean(glucose_value_numeric),
    median_glucose = median(glucose_value_numeric),
    sd_glucose     = sd(glucose_value_numeric),
    cv_glucose     = sd_glucose / mean_glucose * 100,
    pct_70_180     = mean(glucose_value_numeric >= 70 & glucose_value_numeric <= 180) * 100,
    pct_above_180  = mean(glucose_value_numeric > 180) * 100,
    pct_above_250  = mean(glucose_value_numeric > 250) * 100,
    pct_below_70   = mean(glucose_value_numeric < 70) * 100,
    pct_below_54   = mean(glucose_value_numeric < 54) * 100
  )

cgm_1029_participant_summary %>% knitr::kable(digits = 1)
```


##### Participant-Day Summary
```{r}
cgm_1029_day_summary <- cgm_1029_valid_days_readings %>%
  mutate(period = if_else(hour_local >= 6, "daytime", "nighttime")) %>%
  group_by(person_id, date_local) %>%
  summarize(
    n_readings         = n(),
    pct_expected        = n_readings / 288 * 100,
    mean_glucose        = mean(glucose_value_numeric),
    sd_glucose           = sd(glucose_value_numeric),
    pct_70_180           = mean(glucose_value_numeric >= 70 & glucose_value_numeric <= 180) * 100,
    pct_above_180         = mean(glucose_value_numeric > 180) * 100,
    pct_below_70           = mean(glucose_value_numeric < 70) * 100,
    n_daytime               = sum(period == "daytime"),
    n_nighttime              = sum(period == "nighttime"),
    mean_glucose_daytime      = mean(glucose_value_numeric[period == "daytime"]),
    mean_glucose_nighttime     = mean(glucose_value_numeric[period == "nighttime"]),
    sd_glucose_daytime          = sd(glucose_value_numeric[period == "daytime"]),
    sd_glucose_nighttime         = sd(glucose_value_numeric[period == "nighttime"]),
    .groups = "drop"
  )

cgm_1029_day_summary %>% knitr::kable(digits = 1)
```
