---
title: "Week 3: Sleep Activity Processing"
author: "Eric Xie"
format:
  html:
    toc: true
    toc-location: left
---

## Load Packages
```{r}
library(tidyverse)
library(knitr)
library(readr)
library(jsonlite)
library(kableExtra)
```

## Person-Night Sleep Table
```{r}
sleep_path <- ("data/mini/sleep/garmin_vivosmart5/1029/1029_sleep.json")
sleep_json <- fromJSON(sleep_path)

stages <- sleep_json$body$sleep

stages <- stages |>
  unnest_wider(effective_time_frame) |>
  unnest_wider(time_interval)

sleep_tbl <- stages |>
  mutate(
    start = ymd_hms(start_date_time, tz = "UTC"),
    end   = ymd_hms(end_date_time, tz = "UTC"),
    duration_min = as.numeric(difftime(end, start, units = "mins")),
    # 18 as the cutoff hour for the next day
    sleep_night = as_date(if_else(hour(start) < 18, start - days(1), start))
  ) |>
  group_by(participant_id = 1029, sleep_night) |>
  summarise(
    bedtime = min(start), 
    waketime = max(end),
    total_sleep_time_min = sum(duration_min[sleep_stage_state != "awake"]),
    time_in_bed_min = as.numeric(difftime(max(end), 
                                          min(start), 
                                          units = "mins")),
    sleep_efficiency_pct  = total_sleep_time_min / time_in_bed_min * 100,
    pct_light = sum(duration_min[sleep_stage_state == "light"]) / sum(duration_min) * 100,
    pct_deep = sum(duration_min[sleep_stage_state == "deep"]) / sum(duration_min) * 100,
    pct_rem = sum(duration_min[sleep_stage_state == "rem"]) / sum(duration_min) * 100,
    pct_awake = sum(duration_min[sleep_stage_state == "awake"]) / sum(duration_min) * 100,
    ) |>
  arrange(sleep_night)

kable(sleep_tbl, digits = 2)
```

##### Analysis
- On 9/11, 9/13, and 9/16, Participant 1029 had very good sleep quality (sleep efficiency 100%)
- However, on 9/16, Participant 1029 had less than 5 hours (300 minutes) of sleep time, much less than other nights.
- On 9/9, the participant was almost 30% awake of the night time. 

## Daily Activity Table

```{r}
pa_path <- ("data/mini/physical_activity/garmin_vivosmart5/1029/1029_activity.json")
pa_json <- fromJSON(pa_path)

pac_path <- ("data/mini/physical_activity_calorie/garmin_vivosmart5/1029/1029_calorie.json")
pac_json <- fromJSON(pac_path)

hr_path <- ("data/mini/heart_rate/garmin_vivosmart5/1029/1029_heartrate.json")
hr_json <- fromJSON(hr_path)

# steps
steps_raw <- pa_json$body$activity |>
  unnest_wider(base_movement_quantity, names_sep = "_") |>
  unnest_wider(effective_time_frame) |>
  unnest_wider(time_interval)

steps_daily <- steps_raw |>
  mutate(
    start = ymd_hms(start_date_time, tz = "UTC"),
    date  = as_date(start),
    steps = as.numeric(base_movement_quantity_value)
  ) |>
  group_by(date) |>
  summarise(total_steps = sum(steps, na.rm = TRUE))

# calories
cal_raw <- pac_json$body$activity |>
  unnest_wider(calories_value, names_sep = "_") |>
  unnest_wider(effective_time_frame)

cal_daily <- cal_raw |>
  mutate(
    ts   = ymd_hms(date_time, tz = "UTC"),
    date = as_date(ts)
  ) |>
  group_by(date) |>
  summarise(active_calories = sum(calories_value_value, na.rm = TRUE), .groups = "drop")

# heart rate
hr_raw <- hr_json$body$heart_rate |>
  unnest_wider(heart_rate, names_sep = "_") |>
  unnest_wider(effective_time_frame)

hr_daily <- hr_raw |>
  mutate(
    ts   = ymd_hms(date_time, tz = "UTC"),
    date = as_date(ts)
  ) |>
  filter(heart_rate_value != 0) |>
  group_by(date) |>
  summarise(average_heartrate_bpm = mean(heart_rate_value, na.rm = TRUE))

# join into one table
daily_activity_tbl <- steps_daily |>
  full_join(cal_daily, by = "date") |>
  full_join(hr_daily,  by = "date") |>
  mutate(participant_id = pa_json$header$user_id, .before = 1) |>
  arrange(date)

kable(daily_activity_tbl, digits = 1)
```

##### Analysis

- On 9/19, the participant had 0 total steps recorded (likely missing data)
- Average heart rate seems to be relatively constant throughout the study (hovering around the 60s and 70s)

## Sleep Duration Plot
```{r}
ggplot(sleep_tbl, aes(x = sleep_night, y = total_sleep_time_min)) +
  geom_col()
```

- 9/12 has missing data on sleep time

## Daily Steps Plot
```{r}
ggplot(daily_activity_tbl, aes(x = date, y = total_steps)) +
  geom_col()
```

- 9/7 is a partial day (starting from 5:43 pm), so the recorded steps is lower compared to the other days

## CGM During Sleep Plot
```{r}
# load glucose data
glucose_json <- fromJSON("data/mini/dexcom_g6/1029/1029_DEX.json")

glucose_raw <- glucose_json$body$cgm |>
  unnest_wider(blood_glucose, names_sep = "_") |>
  unnest_wider(effective_time_frame) |>
  unnest_wider(time_interval)

glucose_tbl <- glucose_raw |>
  mutate(reading_time = ymd_hms(start_date_time, tz = "UTC")) |>
  select(reading_time, glucose_mg_dl = blood_glucose_value)

# join
glucose_during_sleep <- glucose_tbl |>
  cross_join(sleep_tbl |> select(participant_id, sleep_night, bedtime, waketime)) |>
  filter(reading_time >= bedtime, reading_time <= waketime) |>
  select(participant_id, sleep_night, reading_time, glucose_mg_dl)

# plot
ggplot(glucose_during_sleep, aes(x = reading_time, y = glucose_mg_dl)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ sleep_night, scales = "free_x", ncol = 3) +
  labs(
    title = "CGM During Sleep — Participant 1029",
    x = "Time",
    y = "Glucose (mg/dL)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```

##### Analysis

- From nights 9/11 to 9/16, the graph was relatively constant (fluctuations in glucose is small)
- 9/10 has the biggest glucose fluctuation




