---
title: "Week 4: Integrating Sleep, Activity, and Glucose"
author: "Eric Xie"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
engine: knitr
---

## Load Packages
```{r}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
library(janitor)
library(jsonlite)
```

## Reviewing Week 2 Solution Keys for Participant-Day CGM Table

### Analysis Parameters
```{r}
expected_readings_per_day <- 288
min_readings_valid_day <- 202
glucose_lower_bound <- 40
glucose_upper_bound <- 400
daytime_start_hour <- 6
```

### Dataset Path
```{r}
# dataset path
dataset_version <- "v3-mini"

data_dir <- here("data/mini")
```

### Read Participants Manifest Data
```{r}
participants <- read_tsv(
  file.path(
    data_dir,
    "participants_mini.tsv"
  ),
  show_col_types = FALSE
) |>
  clean_names()
```

- clean_names() is a janitor function that standardizes column names into a consistent, lowercase, snake_case format

### Cohort Files
```{r}
# cohort files
dexcom_dir <- file.path(
  data_dir,
  "dexcom_g6"
)

# define a vector for the cgm cohort
cgm_cohort <- participants |>
  filter(wearable_blood_glucose) |>
  pull(person_id)

participant_dirs <- list.dirs(dexcom_dir, recursive = FALSE, full.names = TRUE)

cohort_files <- tibble(dir_path = participant_dirs) |>
  mutate(
    folder_name = basename(dir_path),
    person_id = suppressWarnings(as.numeric(str_remove(folder_name, "AIREADI-"))),
    json_path = map_chr(dir_path, function(one_dir) {
      found <- list.files(one_dir, pattern = "\\.json$", full.names = TRUE)
      if (length(found) >= 1) found[1] else NA_character_
    })
  ) |>
  filter(person_id %in% cgm_cohort, !is.na(json_path))
```

- cgm_cohort filtered for participants with wearable_blood_glucose as true, and turned the data frame into a vector with person ids
- cohort_files added folder name, person id, and json paths to the table with dir_path and ensured consistent formatting

### Timezone Tables
```{r}
participant_sites <- participants |>
  select(person_id, clinical_site)

tz_map <- tribble(
  ~timezone_label_json, ~iana_zone,           ~expected_region,
  "pst",                "America/Los_Angeles", "Pacific",
  "pdt",                "America/Los_Angeles", "Pacific",
  "mst",                "America/Denver",      "Mountain",
  "mdt",                "America/Denver",      "Mountain",
  "cst",                "America/Chicago",     "Central",
  "cdt",                "America/Chicago",     "Central",
  "est",                "America/New_York",    "Eastern",
  "edt",                "America/New_York",    "Eastern"
)

site_region_map <- tribble(
  ~clinical_site, ~site_region,
  "UW",           "Pacific",
  "UCSD",         "Pacific",
  "UAB",          "Central"
)
```

### Function for clean_one_cgm
```{r}
clean_one_cgm <- function(json_path) {
  
  # flattening into tables is quicker than iterating through nested objects
  raw <- fromJSON(
    json_path,
    flatten = TRUE
  )
  
  person_id_header <- raw$header$patient_id

  pid <- as.numeric(
    str_remove(person_id_header, "AIREADI-")
  )

  tz_label <- str_to_lower(raw$header$timezone)

  cgm_raw <- as_tibble(raw$body$cgm)

  # Step 1: keep estimated glucose value events only.
  # Step 2: timestamps end in "Z", so they are parsed as UTC.
  #
  # as.character() is applied to the glucose value before as.numeric()
  # because Dexcom writes "Low" and "High" as sentinel strings. This makes
  # the column character in some files and integer in others, and the two
  # types cannot be combined across participants without an explicit cast.

  egv <- cgm_raw |>
    filter(event_type == "EGV") |>
    transmute(
      person_id = pid,
      event_type,
      timezone_label_json = tz_label,
      time_utc = ymd_hms(
        effective_time_frame.time_interval.start_date_time,
        tz = "UTC",
        quiet = TRUE
      ),
      glucose_value_original = as.character(blood_glucose.value),
      glucose_value_numeric = suppressWarnings(
        as.numeric(glucose_value_original)
      ),
      glucose_unit = blood_glucose.unit
    ) # each row represents one glucose reading

  n_raw_egv <- nrow(egv) # counts how many rows are in egv data frame

  # Step 3: infer the regional timezone from the JSON label and check it
  # against the clinical site.
  #
  # The JSON header says "pst" and the clinical site is UW.
  # We interpret this as US Pacific Time and use "America/Los_Angeles"
  # for its timezone and daylight-saving rules. R determines whether
  # each individual timestamp is PST or PDT.
  #
  # A label that cannot be mapped at all means we cannot localize the
  # readings, so the participant is flagged and returns no clean readings.
  # A label that maps successfully but disagrees with the clinical site is
  # also flagged for human review, rather than being silently overridden.

  tz_match <- tz_map |>
    filter(timezone_label_json == tz_label)

  site_value <- participant_sites |>
    filter(person_id == pid) |>
    pull(clinical_site)

  site_match <- site_region_map |>
    filter(clinical_site %in% site_value)

  iana_zone <- if (nrow(tz_match) == 1) {
    tz_match$iana_zone
  } else {
    NA_character_
  }

  expected_region <- if (nrow(tz_match) == 1) {
    tz_match$expected_region
  } else {
    NA_character_
  }

  site_region <- if (nrow(site_match) == 1) {
    site_match$site_region
  } else {
    NA_character_
  }

  tz_site_consistent <- !is.na(expected_region) &
    !is.na(site_region) &
    expected_region == site_region

  tz_qc <- tibble(
    person_id = pid,
    timezone_label_json = tz_label,
    iana_zone = iana_zone,
    clinical_site = if (length(site_value) == 1) {
      site_value
    } else {
      NA_character_
    },
    expected_region = expected_region,
    site_region = site_region,
    tz_site_consistent = tz_site_consistent,
    tz_flag_for_review = is.na(iana_zone) | !tz_site_consistent
  )
  
  # placeholder list in case of errors
  empty_result <- list(
    reading_level = tibble(),
    day_counts = tibble(),
    dst_notes = tibble(),
    tz_qc = tz_qc,
    qc_summary = tibble(
      person_id = pid,
      raw_egv_readings = n_raw_egv,
      missing_timestamps = sum(is.na(egv$time_utc)),
      missing_glucose = sum(is.na(egv$glucose_value_numeric)),
      duplicate_timestamps = NA_integer_,
      conflicting_duplicates = NA_integer_,
      nonstandard_units = NA_integer_,
      unusual_values = NA_integer_,
      cleaned_readings = 0L,
      valid_days = 0L,
      excluded_duplicate_readings = NA_integer_,
      excluded_missing_glucose = NA_integer_,
      excluded_units = NA_integer_,
      excluded_unusual = NA_integer_
    )
  )

  if (is.na(iana_zone) || n_raw_egv == 0) {
    return(empty_result)
  }

  # Step 4: convert to local time and derive the local calendar variables.
  # timezone_abbreviation is taken per reading rather than from the header,
  # so a participant recorded across a transition shows both PDT and PST.

  egv <- egv |>
    mutate(
      time_local = with_tz(time_utc, tzone = iana_zone),
      date_local = as_date(time_local),
      hour_local = hour(time_local),
      timezone_abbreviation = format(time_local, "%Z"),
      is_dst = dst(time_local)
    )

  # Step 6: count missing timestamps and missing glucose values.
  # "Low" and "High" become NA on conversion and are counted here.

  n_missing_timestamps <- sum(is.na(egv$time_utc))

  n_missing_glucose <- sum(is.na(egv$glucose_value_numeric))

  # Step 7: a duplicate is the same participant ID and UTC timestamp.
  # Duplicates that agree on glucose collapse to a single copy.
  # Duplicates that disagree are flagged and excluded entirely, because
  # there is no basis for choosing between the conflicting values.
  #
  # This rule is from Shubho's work. It is
  # adopted here in preference to keeping the first copy, because the
  # cohort does contain conflicting duplicates.

  duplicate_groups <- egv |>
    filter(!is.na(time_utc)) |>
    group_by(person_id, time_utc) |>
    summarise(
      number_of_rows = n(),
      number_of_distinct_values = n_distinct(glucose_value_numeric),
      .groups = "drop"
    ) |>
    filter(number_of_rows > 1)

  conflicting_keys <- duplicate_groups |>
    filter(number_of_distinct_values > 1) |>
    select(person_id, time_utc)

  n_duplicate_timestamps <- nrow(duplicate_groups)

  n_conflicting_duplicates <- nrow(conflicting_keys)

  egv_timed <- egv |>
    filter(!is.na(time_utc))

  egv_dedup <- egv_timed |>
    anti_join(
      conflicting_keys,
      by = c("person_id", "time_utc")
    ) |>
    distinct(person_id, time_utc, .keep_all = TRUE)

  # Duplicate handling removes whole rows, and one conflicting group of two
  # rows removes two readings. This row count, not the group count, is what
  # the cohort funnel below needs in order to reconcile exactly.

  n_duplicate_readings_removed <- nrow(egv_timed) - nrow(egv_dedup)

  # Step 8: confirm that glucose is recorded in mg/dL. Any other unit is
  # flagged and excluded, because the 40-400 range check below and every
  # summary in this notebook assume mg/dL.

  egv_dedup <- egv_dedup |>
    mutate(
      unit_flag = is.na(glucose_unit) | glucose_unit != "mg/dL"
    )

  # Step 9: flag values below 40 or above 400 mg/dL as unusual and exclude
  # them. This is a data-processing rule, not a clinical definition.

  egv_dedup <- egv_dedup |>
    mutate(
      unusual_value_flag = !is.na(glucose_value_numeric) &
        (glucose_value_numeric < glucose_lower_bound |
           glucose_value_numeric > glucose_upper_bound)
    )

  n_nonstandard_units <- sum(egv_dedup$unit_flag)

  n_unusual <- sum(egv_dedup$unusual_value_flag)

  # The three exclusion rules above can apply to the same reading, so the
  # cohort funnel needs counts that charge each excluded reading to exactly
  # one rule. These cascade in the same order the filter below applies.

  n_excluded_missing_glucose <- sum(
    is.na(egv_dedup$glucose_value_numeric)
  )

  n_excluded_units <- sum(
    !is.na(egv_dedup$glucose_value_numeric) &
      egv_dedup$unit_flag
  )

  n_excluded_unusual <- sum(
    !is.na(egv_dedup$glucose_value_numeric) &
      !egv_dedup$unit_flag &
      egv_dedup$unusual_value_flag
  )

  # Step 10: the clean reading-level table. One row is one unique, usable
  # glucose reading. The raw objects above are left unchanged.

  reading_level <- egv_dedup |>
    filter(
      !is.na(glucose_value_numeric),
      !unit_flag,
      !unusual_value_flag
    ) |>
    select(
      person_id,
      event_type,
      time_utc,
      time_local,
      date_local,
      hour_local,
      timezone_abbreviation,
      timezone_label_json,
      is_dst,
      glucose_value_original,
      glucose_value_numeric,
      glucose_unit
    )

  # Step 11: count cleaned readings per local calendar day and mark a
  # participant-day valid at 202 or more of the expected 288 readings.

  day_counts <- reading_level |>
    count(
      person_id,
      date_local,
      name = "n_readings"
    ) |>
    mutate(
      valid_day = n_readings >= min_readings_valid_day
    )

  # Step 12: flag local dates that span a daylight-saving-time transition.
  # The valid-day threshold is deliberately left unchanged.

  # Abbreviations are ordered by time rather than sorted alphabetically,
  # so that a spring-forward date reads "PST to PDT" and a fall-back date
  # reads "PDT to PST". Sorting would label both the same way.
  #
  # The true length of a local day is the distance between consecutive
  # local midnights in the participant's own zone, which is 23 hours on a
  # spring-forward date and 25 hours on a fall-back date. Midnight itself
  # always exists, because US transitions occur at 2:00 AM local time.

  dst_notes <- reading_level |>
    group_by(person_id, date_local) |>
    summarise(
      abbreviations = paste(
        unique(timezone_abbreviation[order(time_local)]),
        collapse = " to "
      ),
      number_of_abbreviations = n_distinct(timezone_abbreviation),
      .groups = "drop"
    ) |>
    filter(number_of_abbreviations > 1) |>
    mutate(
      day_length_hours = as.numeric(
        difftime(
          force_tz(as_datetime(date_local + 1), tzone = iana_zone),
          force_tz(as_datetime(date_local), tzone = iana_zone),
          units = "hours"
        )
      ),
      transition = if_else(
        day_length_hours < 24,
        "Spring forward",
        "Fall back"
      ),
      max_possible_readings = day_length_hours * 12,
      qc_note = paste0(
        transition,
        " transition (",
        abbreviations,
        "). The local day is ",
        day_length_hours,
        " hours long, so at most ",
        max_possible_readings,
        " readings are possible instead of ",
        expected_readings_per_day,
        "."
      )
    )

  day_counts <- day_counts |>
    mutate(
      crosses_dst = date_local %in% dst_notes$date_local
    )

  # The first block of columns is the quality-control summary requested for
  # Table 1, and its schema and column names come from Eric's work.
  # The excluded_* columns are the mutually exclusive cascade
  # used by the cohort funnel, and are not displayed in Table 1.

  qc_summary <- tibble(
    person_id = pid,
    raw_egv_readings = n_raw_egv,
    missing_timestamps = n_missing_timestamps,
    missing_glucose = n_missing_glucose,
    duplicate_timestamps = n_duplicate_timestamps,
    conflicting_duplicates = n_conflicting_duplicates,
    nonstandard_units = n_nonstandard_units,
    unusual_values = n_unusual,
    cleaned_readings = nrow(reading_level),
    valid_days = sum(day_counts$valid_day),
    excluded_duplicate_readings = n_duplicate_readings_removed,
    excluded_missing_glucose = n_excluded_missing_glucose,
    excluded_units = n_excluded_units,
    excluded_unusual = n_excluded_unusual
  )

  list(
    reading_level = reading_level,
    day_counts = day_counts,
    dst_notes = dst_notes,
    tz_qc = tz_qc,
    qc_summary = qc_summary
  )
}
```

### Create Results
```{r}
safe_clean_one_cgm <- possibly(clean_one_cgm, otherwise = NULL)

results_all <- map(
  cohort_files$json_path,
  safe_clean_one_cgm
)

reading_level_all <- results_all |>
  map_dfr("reading_level")

day_counts_all <- results_all |>
  map_dfr("day_counts")

valid_readings_all <- reading_level_all |>
  inner_join(
    day_counts_all |>
      filter(valid_day) |>
      select(person_id, date_local),
    by = c("person_id", "date_local")
  )

participant_day_summary <- valid_readings_all |>
  mutate(
    period = if_else(
      hour_local >= daytime_start_hour,
      "daytime",
      "nighttime"
    )
  ) |>
  group_by(person_id, date_local) |>
  summarise(
    n_readings = n(),
    pct_expected = n() / expected_readings_per_day * 100,
    mean_glucose = mean(glucose_value_numeric),
    sd_glucose = sd(glucose_value_numeric),
    pct_70_180 = mean(
      glucose_value_numeric >= 70 & glucose_value_numeric <= 180
    ) * 100,
    pct_above_180 = mean(glucose_value_numeric > 180) * 100,
    pct_below_70 = mean(glucose_value_numeric < 70) * 100,
    n_daytime = sum(period == "daytime"),
    n_nighttime = sum(period == "nighttime"),
    mean_glucose_daytime = mean(
      glucose_value_numeric[period == "daytime"]
    ),
    mean_glucose_nighttime = mean(
      glucose_value_numeric[period == "nighttime"]
    ),
    sd_glucose_daytime = sd(
      glucose_value_numeric[period == "daytime"]
    ),
    sd_glucose_nighttime = sd(
      glucose_value_numeric[period == "nighttime"]
    ),
    .groups = "drop"
  ) |>
  mutate(
    across(
      where(is.numeric),
      ~ if_else(is.nan(.x), NA_real_, .x)
    )
  ) |>
  left_join(
    day_counts_all |>
      select(person_id, date_local, crosses_dst),
    by = c("person_id", "date_local")
  )

participant_day_summary |>
  kable(
    digits = 1,
    col.names = c(
      "Person ID",
      "Local date",
      "Readings",
      "Expected (%)",
      "Mean",
      "SD",
      "70-180 (%)",
      ">180 (%)",
      "<70 (%)",
      "Day n",
      "Night n",
      "Day mean",
      "Night mean",
      "Day SD",
      "Night SD",
      "Crosses DST"
    )
  ) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    font_size = 11
  ) |>
  scroll_box(height = "400px")
```

### Save Results
```{r}
dir.create(
  here("practice", "derived"),
  recursive = TRUE,
  showWarnings = FALSE
)

write_csv(
  participant_day_summary,
  here(
    "practice",
    "derived",
    "week2_participant_day_cgm.csv"
  ),
  na = ""
)
```

### Analysis

- `raw`: the parsed Json file, a nested list that was flattened into a table for easier iterating.
- `egv`: filtered to valid glucose readings through `event_type == "EGV"`, where each row contains one raw estimated glucose reading. 
- **Timezone Resolution**: matching the JSON header's timezone label against `tz_map`, and the participant's clinical site against `site_region_map` with key `person_id`. Each row contains one participant's timezone quality-control record.`iana_zone` is resolved to "NA" if the label doesn't match to any defined iana zone.
- `egv_dedup`: drops duplicate timestamps by reference `(person_id, time_utc)`. Flags any inconsistent units and unusual glucose values.
- `reading_level`: the final clean reading-level table, where each row is one unique glucose reading.
- `day_counts`: built by `count(person_id, date_local)` on `reading_level`. Each row is one unique participant day. A day is only valid with more than 202 readings of the expected 288. 
- `dst_notes`: DST-Transition days by participant, each row containing a participant-day that crossed the DST boundary.
- `participant_day_summary`: final analysis table, with each row equal to one participant-day. Key = `(person_id, time_utc)`.




